A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-12 jour/jour.tex
more

%\begin{lemma}
%\label{lem:mu-norm-interp}
%If the $\mu$-term $t$ is normal and $u$ is a subterm of $t$, then $\expand{t}{u}$ is normal.
%\end{lemma}
% TODO:
%\begin{proof}
%\qed
%\end{proof}

@@

The decision procedure was presented at a high level of abstraction, omitting
quite a few details. This section describes the main aspects of the
implementation within the SMT solver CVC4:\ the integration into CDCL($T$)
\cite{ganzinger-et-al-2004}, the extension to quantified formulas, and some of
the optimizations.

%%% @ANDY: I'm trying without \paragraph's, to save space but also to make the
%%% section look more like a whole than a random list of stuff. Revert if you
%%% disagree.
\subsection{Theory Solver for $\thD$}
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

\begin{figure}[t!]
\normalsize
\centering
\begin{enumerate}
\item Let $U$ be the datatype equivalence classes of $\Ec$.
\item Repeat until $U \neq \emptyset$.
\begin{itemize}
\item[\ ] For some $\ec{t} \in U$, $\mathrm{traverse}( \ec{t}, U, \emptyset )$
\end{itemize}
\end{enumerate}

\flushleft $\mathrm{traverse}( \ec{t}, U, P )$:
\begin{enumerate}
\item If $\ec{t} \in P$, apply \rn{Acyclic} to $t$.
\item If $\ec{t} \in U$,
\begin{enumerate}
\item If $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$, then for each $j = 1, \ldots k$,
\begin{itemize}
\item[\ ] $\mathrm{traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} )$
\end{itemize}
\item $U := U \textbackslash \{ \ec{t} \}$.
\end{enumerate}
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking \rn{Acyclic}}
\label{fig:acyclic-algo}
\end{figure}

\begin{figure}[t!]
\normalsize
\centering
\begin{enumerate}
\item Let $\mathcal{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{t} =_{\mathcal{U}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $t\typ{\delta}$, $u\typ{\delta}$ where $\delta \in \Codata$, and $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$.
\end{itemize}
\item Repeat until $\mathcal{U}$ is unchanged
\begin{enumerate}
\item $\mathcal{U}' := \emptyset$
\item for each $U_i \in \mathcal{U}$:
\begin{itemize}
\item[\ ] Let $\mathcal{U}_i$ be a partition of $U_i$ such that $\ec{t} =_{\mathcal{U}_i} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcal{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[\ ] $\mathcal{U}' := \mathcal{U}' \cup \mathcal{U}_i$.
\end{itemize}
\item $\mathcal{U} := \mathcal{U}'$
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcal{U}$, apply \rn{Unique} to $t,u$.
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, we use the algorithms in Figures~\ref{fig:acyclic-algo} and~\ref{fig:uniqueness-algo}
for determining whether the rules \rn{Acyclic} and \rn{Unique} apply to our set $\Ec$.

For \rn{Acyclic}, Figure~\ref{fig:acyclic-algo} considers the set $U$ of all datatype equivalence classes of $\Ec$.
We choose an arbitrary $\ec{t}$ in $U$ and call the recursive subprocedure $\mathrm{traverse}$,
which takes as input the current equivalence class we are processing, the set of equivalence classes $U$ we have yet to process,
and the set of equivalence classes $P$ we are currently processing.
If a call to $\mathrm{traverse}( \ec{t}, U, P )$ is such that $\ec{t} \in P$,
we know that the rule \rn{Acyclic} applies to $t$.
Otherwise, if we have yet to process $\ec{t}$, as indicated by the condition $\ec{t} \in U$,
then if $\ec{t}$ contains a constructor term $\const{C}( t_1, \ldots, t_k )$, 
then we recursively call $\mathrm{traverse}$ on each of $\ec{t_1}, \ldots, \ec{t_n}$.
If this succeeds, we remove $\ec{t}$ from $U$.

For \rn{Unique}, Figure~\ref{fig:uniqueness-algo} considers a partition $\mathcal{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcal{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcal{U}$.
We then refine $\mathcal{U}$ by 
constructing a partition $\mathcal{U}_i$ of each $U_i \in \mathcal{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcal{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcal{U}$ for each $j=1, \ldots, k$.
Let $\mathcal{U'}$ be the union of each of these partitions.
We update $\mathcal{U}$ to $\mathcal{U'}$.
This refinement is repeated until $\mathcal{U}$ is left unchanged.
If the resulting partition $\mathcal{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is similar to the classic algorithm for minimizing deterministic finite automata (DFA). \textbf{maybe cite?}

@@

\subsection{Extension to Quantified Formulas}
While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to %quantified
formulas involving datatypes and codatatypes.

>>>>

%\begin{lemma}
%\label{lem:mu-norm-interp}
%If the $\mu$-term $t$ is normal and $u$ is a subterm of $t$, then $\expand{t}{u}$ is normal.
%\end{lemma}
%\begin{proof}
%\qed
%\end{proof}

@@

The decision procedure was presented at a high level of abstraction, omitting
quite a few details. This section describes the main aspects of the
implementation within the SMT solver CVC4:\ the integration of the
procedure into CDCL($T$)
\cite{ganzinger-et-al-2004}, the construction of models with $\mu$-terms,
and the extension of the procedure to quantified formulas.
%, and some of the optimizations.

\subsection{A Theory Solver for $\thD$}

The decision procedure is implemented as a \relax{theory solver} of CVC4---that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of
\begin{itemize}
\item the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
\item additional (dis)equalities between variables of types not belonging to
$T_i$.
\end{itemize}
Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

\begin{figure}[t!]
\normalsize
\centering
\flushleft $\rn{Acyclic}(t)$:

\vspace*{-\topsep}

\begin{enumerate}
\item Let $U$ be the datatype equivalence classes of $\Ec$.
\item Repeat until $U \neq \emptyset$:
\begin{itemize}
\item[\ ] For some $\ec{t} \in U$, $\rn{Traverse}( \ec{t}, U, \emptyset )$.
\end{itemize}
\end{enumerate}

\flushleft $\rn{Traverse}( \ec{t}, U, P )$:

\vspace*{-\topsep}

\begin{enumerate}
\item If $\ec{t} \in P$, $\rn{Acyclic}(t)$.
\item If $\ec{t} \in U$:
\begin{enumerate}
\item[2.1.] If $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$, then for each $j = 1, \ldots k$:
\begin{itemize}
\item[2.1.1.] $\rn{Traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} )$.
\end{itemize}
\item[2.2.] $U := U \textbackslash \{ \ec{t} \}$.
\end{enumerate}
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Acyclic}}
\label{fig:acyclic-algo}
\end{figure}

\begin{figure}[t!]
\normalsize
\centering

\flushleft $\rn{Unique}(t, u)$:

\vspace*{-\topsep}

\begin{enumerate}
\item Let $\mathcalx{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{t} =_{\mathcalx{U}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
  $t$, $u$ have type $\delta \in \Codata$,
  $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$,
  and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$.

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset$.
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_i$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_i} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_i$.
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'$.
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U}$, $\rn{Unique}(t,u)$.
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, it relies on the algorithms in Figs.~\ref{fig:acyclic-algo} and~\ref{fig:uniqueness-algo}
for determining whether the rules \rn{Acyclic} and \rn{Unique} apply to the set $\Ec$.

For \rn{Acyclic}, Fig.~\ref{fig:acyclic-algo} considers the set $U$ of all datatype equivalence classes of $\Ec$.
We choose an arbitrary $\ec{t}$ in $U$ and call the recursive subprocedure $\rn{Traverse}$,
which takes as input the current equivalence class we are processing, the set of equivalence classes $U$ we have yet to process,
and the set of equivalence classes $P$ we are currently processing.
If a call to $\rn{Traverse}( \ec{t}, U, P )$ is such that $\ec{t} \in P$,
we know that the rule \rn{Acyclic} applies to $t$.
Otherwise, if we have yet to process $\ec{t}$, as indicated by the condition $\ec{t} \in U$,
then if $\ec{t}$ contains a constructor term $\const{C}( t_1, \ldots, t_k )$, 
then we recursively call $\rn{Traverse}$ on each of $\ec{t_1}, \ldots, \ec{t_n}$.
If this succeeds, we remove $\ec{t}$ from $U$.

For \rn{Unique}, Fig.~\ref{fig:uniqueness-algo} considers a partition $\mathcalx{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcalx{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcalx{U}$.
We then refine $\mathcalx{U}$ by 
constructing a partition $\mathcalx{U}_i$ of each $U_i \in \mathcalx{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcalx{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcalx{U}$ for each $j=1, \ldots, k$.
Let $\mathcalx{U'}$ be the union of each of these partitions.
We update $\mathcalx{U}$ to $\mathcalx{U'}$.
This refinement is repeated until $\mathcalx{U}$ is left unchanged.
If the resulting partition $\mathcalx{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is analogous to Hopcroft's algorithm for minimizing
deterministic finite automata \cite{hopcroft-1971}---terms correspond to
states, argument positions $1,\ldots,k$ correspond to input symbols, and the top constructor symbol
of a term generalizes the accepting/rejecting status of a state.

@@

\subsection{Extension to Quantified Formulas}

While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to %quantified
formulas involving datatypes and codatatypes.
